---
title: "Assignment 2: Coding Basics"
author: "Kamil Orozco"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. #assigning the sequence
FirstRSeq <- seq(1, 100, 4) 

#2. #summary statistics
MedFirstSeq <- median(FirstRSeq) 
MeanFirstSeq <- mean(FirstRSeq) 

#3. #conditional formatting
MeanFirstSeq > MedFirstSeq



```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

```{r}
# data frame components
Name <- c("Meliodas", "Escanor", "Merlin", "Gowther")
Scoring <- c(72, 26, 64, 89)
# Which scores are >= 50?
Grade <- c(TRUE, FALSE, TRUE, TRUE)



```


6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#forming data frame
StudentScoreSheet <- data.frame(Name, Scoring, Grade)
StudentScoreSheet
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: It is different from a matrix because a matrix can house a single class of data; data frames can contain multiple classes of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
# creating an if/else statement on pre-made Scoring vector
Test.Scores <- function(GradeList){
for(val in Scoring){
  if(val >=50){
    print (TRUE)
  } else {
    print (FALSE)}
}  
}  

 
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The "ifelse" function returns value data instead of the printed true false statements that the "if" "else" function provides. This way, we are able to see which students got a certain score greater than or equal to 50 in the syntax form rather than the quantifiable form (like a percentage).
